Associative Data Structures for Website Filtering

### 2014-04-17 paper/paper.tex
Reworked conclusion and reflection.

    \subsection*{Data Structures}

@@

    \subsubsection*{Data}
    For our data, we used datasets containing $2^x$ elements with $x$ an integer such that $8 < x <
    21$. Thus we have twelve different dataset sizes. We took the system's dictionary file and the
    list of top level domains from Wikipedia and combined these to generate a list of items in the
    form \texttt{www.\$word.\$tld}, where \texttt{\$word} is one of words from the the dictionary
    and \texttt{\$tld} is a top level domain.  Only words consisting of the lowercase letters
    \texttt{a} through \texttt{z} were used. Constraining the words to consist of 8 to 12 characters
    and limiting the top-level domains to exactly two letters, all possible combinations were
    generated.  For the full paths, a number of domains were selected and each was appended a random
    alphanumeric string of eighty characters.  Finally, for the third type of key, an IP address, we
    used 32 bit integers. The exact implementation can be viewed in the sourcecode\footnote{ref to
    apendix or code dump}.

    \subsection*{Measurements}
    For our time measurements, we used libstdc++'s implementation of the standard
    \texttt{std::chrono::high\_resolution\_clock} class.

@@

    \section{Conclusions}
    Our main point of research was researching which associative data structure performed best on a
    various number of insertions, in terms of the speed of insertions and look-ups, and memory
    usage.  In terms of speed we can state with certainty that the hash table is the superior data
    structure.  It outperforms the other two in almost every category, most significantly in larger
    datastructures where it often takes less than half as much time as the skiplist and less than
    two thirds the time of the binary search tree.  In most cases, the binary search tree is a bit
    slower, whereas the skip list is the slowest structure in virtually every case. The only
    motivation for using a hashmap for this problem appears to be in memory usage, although the
    advantage gained from this is relatively small.

    Our secondary point of research encompassed the effect of key complexity on associative data
    structures.  Our different types of keys were IP adresses, domain names and full path URLs. We
    have seen that as expected, IP adresses as keys provide a significant improvement over the other
    two, and uses less memory as well. Interesting to note is that we found the only advantage of
    the binary search tree over the hashmap here, is that insertions of a small amount of IP
    adresses, up to around $2^13$ insertions, is faster. So when writing an application with very
    few keys of low complexity, and with more insertions than queries, the binary search tree may be
    the best option. Finally, we have found that the domain key type does not outperform the full
    path key type. It is not significantly better if it is better at all.

    \section{Reflection}
    In general, we are content with the results of our research. We were able to provide clear and
    concise answers to our research questions. Even though our research questions may appear broad,
    we have defined a good scope in our research proposal and have virtually not altered this scope
    in our research. One small change is that we also included datasets of size $2^8$ because we saw
    some interesting results in very small sets. In our implementation we have made some significant
    changes because we ran into pratical problems. Some of these required us to change the allocator
    and generate our keys in a slightly different fashion, which we all documented in this paper.

    Since we have found that the hash map is the best associative data structure overal for this
    problem, we could do further research in different kind of hash maps. We might try different
    hash functions, different methods of collision resolution, and perhaps survey different kinds of
    concurrent hash maps.

>>>>

    \subsection{Data Structures}

@@

    \subsection{Data Sources}
    For our data, we used datasets containing $2^x$ elements with $x$ an integer such that $8 < x <
    21$. Thus we have twelve different dataset sizes. We took the system's dictionary file and the
    list of top level domains from Wikipedia and combined these to generate a list of items in the
    form \texttt{www.\$word.\$tld}, where \texttt{\$word} is one of words from the the dictionary
    and \texttt{\$tld} is a top level domain.  Only words consisting of the lowercase letters
    \texttt{a} through \texttt{z} were used. Constraining the words to consist of 8 to 12 characters
    and limiting the top-level domains to exactly two letters, all possible combinations were
    generated.  For the full paths, a number of domains were selected and each was appended a random
    alphanumeric string of eighty characters.  Finally, for the third type of key, an IP address, we
    used 32 bit integers. The exact implementation can be viewed in the sourcecode\footnote{ref to
    apendix or code dump}.

    \subsection{Measurements}
    \label{subsec:measurements}
    For our time measurements, we used libstdc++'s implementation of the standard
    \texttt{std::chrono::high\_resolution\_clock} class.  There is no official documentation about
    the resolution of this clock, but measurements show that it is likely below one microsecond,
    which is sufficient for our purposes.\footnote{\url{http://stackoverflow.com/a/5524138/559931}}

@@

    \section{Conclusions}
    Our main point of research was researching which associative data structure performed best on a
    various number of insertions, in terms of the speed of insertions and queries, and in terms of
    memory necessary per element.  In terms of speed we can state with certainty that the hash table
    is the superior data structure.  It outperforms the other two in almost every category,
    particularly in cases where the number of elements is high, where it is roughly two times faster
    than the skip list and one and a half times faster than the binary search tree.  The binary
    search tree outperforms the hash table when the number of elements is low, and the number of
    insertions is relatively larger than the number of queries.  Whether this situation can occur in
    practice depends on the speed of deletions, which we have not measured; in the given context,
    however, the conditions are unlikely to be achieved.  A skip list may be preferable if memory
    usage is more important than query speeds; given that both insert and query operations take less
    than ten microseconds even in the largest maps we measured, this is likely the case if queries
    and inserts are rare.

    Our secondary point of research encompassed the effect of key complexity on associative data
    structures.  Our different types of keys were IP adresses, domain names and full path URLs.  Our
    measurements confirmed our intuition that IP addresses impose significantly lower costs over the
    other two key types, as well as requiring less memory.  This improvement was more noticable in
    the binary search tree than in the two other systems.  The corresponding expected improvement
    when comparing the domain key type to the full path key type were not observed.

    \section{Reflection}
    In general, we are content with the results of our research. We were able to provide clear and
    concise answers to our research questions. Even though our research questions may appear broad,
    we have defined a good scope in our research proposal and have not altered the overall goal
    of our research.  Errors in the implementation lead us to having to discard measurements of 256
    elements as the outliers were too great to compensate for.  Fortunately, our focus lay primarily
    on the larger datasets and thus this is not significant for our conclusions.  The implementation
    we finally arrived at differs from what we outlined, but the changes are largely to increase the
    accuracy of measurements or the speed at which measurements can be obtained, and have no
    fundamental influence on the results.  The primary difficulty we encountered was the number of
    dimensions in our data: with 117 independent measurements, it took considerable time to perform
    them all, and then to perform statistical analysis on them.

    Knowing the strengths and weaknesses of the various data structures, further research could
    compare subvariants of the hash table and skip list.  In particular, it may be of interest to
    experiment with differen choices of hash function and collision resolution scheme.  Larger
    numbers of elements may also provide interesting challenges, as the number of elements may be
    too great to keep in main memory.  Finally, the situation where a map is accessed by multiple
    threads concurrently provides extra complications that could be considered.
